---
title: "The logratio-normal-multinomial distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lrnm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = F
)
```

```{r setup, message=FALSE}
library(coda.count)
library(coda.base)
library(ggtern)
theme_set(theme_minimal())
```

The logratio-normal-multinomial (LRNM) distribution is the distribution obtained from the compounding of a normal distribution defined on the Simplex using log-ratio coordinates with the multinomial distribution.

We can build a sample of a LRNM as follows:

* Firstly, for given parameters $\mu$ and $Sigma$ and a certain clr-basis, we generate a log-ratio normally distributed sample :

```{r}
set.seed(1)
N = 200
mu = c(0,1)
sigma = matrix(c(1, -0.8, 
                 -0.8, 1), nrow = 2)
B = ilr_basis(3)
P = as.data.frame(rlrnormal(n = N, mu, sigma, B))
H = coordinates(P, B, label = 'ilr')
```

We can visualise either the compositional sample in a Ternary diagram or their coordinates (with respect basis $B$). 

```{r, warning=FALSE, out.width="45%", fig.width=4, fig.height=3.6, fig.show='hold'}
ggtern() +
  geom_point(data = P, aes(x = V1, y = V2, z = V3))
ggplot() +
  geom_point(data = H, aes(x = ilr1, y = ilr2))
```

* After the logratio-normal random variables are generated, they are used to generate a new sample of counts using the multinomial distribution. With the function `rmultinomial()` we can generate that sample.

```{r}
SIZE = 8
Y = as.data.frame(rmultinomial(size = SIZE, p = as.matrix(P)))
```

In `coda.count` we can compound this to generations using the function `rlrnm()`:

```{r}
sample = rlrnm(n = N, size = SIZE, mu = mu, sigma = sigma, probs = TRUE)
Y = as.data.frame(sample)
P = attr(sample, 'probs')
```


To visualise the counting sample we can use a ternary diagram, but observations will only appear in discrete positions. In general, we call the $(K-D)$-simplex lattice the sapce for the $D$-part count data with size $K$.

```{r, include=FALSE}
Yj = as.data.frame(lapply(Y, jitter, factor = 0.5))
ggtern() +
  geom_point(data = Yj, aes(x = V1, y = V2, z = V3), size = 0.5) +
  scale_T_continuous(minor_breaks = NULL, breaks = 0:SIZE/SIZE, labels = 0:SIZE) +
  scale_L_continuous(minor_breaks = NULL, breaks = 0:SIZE/SIZE, labels = 0:SIZE) +
  scale_R_continuous(minor_breaks = NULL, breaks = 0:SIZE/SIZE, labels = 0:SIZE) 
```

```{r, fig.width=6, fig.height=4.8, fig.align='center'}
key = function(x) paste(x, collapse = '-')
tab = table(apply(Y, 1, key))
Yc = Y[duplicated(Y),]
Yc$freq = as.numeric(tab[match(apply(Yc, 1, key), names(tab))])
lbls = paste(' ', 0:SIZE, ' ')
brks = 0:SIZE/SIZE
ggtern() +
  geom_point(data = Yc, aes(x = V1, y = V2, z = V3, col = freq), size = 3) +
  theme(tern.axis.arrow.show = T,
        tern.axis.title.show = F,
        tern.axis.arrow.text = element_text(face = 'bold')) +
  scale_color_gradient(low = 'blue', high = 'red') +
  scale_T_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  scale_L_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  scale_R_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  labs(col = 'Frequency')
```

For given parameters $mu$ and $\Sigma$ we can calculate the probability mass function for each element in a $(8-3)$-simplex lattice:

```{r, fig.width=6, fig.height=4.8, fig.align='center'}
SL3 = simplex_lattice(SIZE, 3)
X = as.data.frame(SL3)
X$p = apply(X, 1, dlrnm, mu, sigma)
ggtern() +
  geom_point(data = X, aes(x = V1, y = V2, z = V3, col = p), size = 3) +
  theme(tern.axis.arrow.show = T,
        tern.axis.title.show = F,
        tern.axis.arrow.text = element_text(face = 'bold')) +
  scale_color_gradient(low = 'blue', high = 'red') +
  scale_T_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  scale_L_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  scale_R_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  labs(col = 'Probability')
```

```{r, eval=FALSE}
SL3 = simplex_lattice(100, 3)
X100 = as.data.frame(SL3)
X100$p = apply(X100, 1, dlrnm, mu, sigma, hermite.order = 100)
X100 = X100[order(X100$p, decreasing = FALSE),]
ggtern() +
  geom_point(data = X100, aes(x = V1, y = V2, z = V3, col = p), size = 1) +
  theme(tern.axis.arrow.show = T,
        tern.axis.title.show = F,
        tern.axis.arrow.text = element_text(face = 'bold')) +
  scale_color_gradient(low = 'blue', high = 'red') +
  labs(col = 'Probability')
```

For the sample `Y` we can

```{r}
fitting = fit_lrnm(Y, probs = TRUE)
```

Expected estimated posterior probabilities

```{r}
P_exp = as.data.frame(fitting$P)
H_exp = coordinates(P_exp)
```

```{r, warning=FALSE, out.width="45%", fig.width=4, fig.height=3.6, fig.show='hold'}
ggtern() +
  geom_point(data = P_exp, aes(x = x1, y = x2, z = x3))
ggplot() +
  geom_point(data = H_exp, aes(x = ilr1, y = ilr2))
```

```{r, fig.width=6, fig.height=4.8, fig.align='center'}
SL3 = simplex_lattice(100, 3)
M = c_obtain_moments_lrnm_hermite(SL3, mu, sigma, ilr_basis(3), order = 10)
H1 = as.data.frame(t(M[[1]]))
M1 = composition(H1)
ellipse = function(mu, sigma, p){
  s = -2 * log(1 - p);
  ev = eigen(sigma * s)
  t_ = seq(0, 2 * pi, length.out = 500)
  a = mu + t((t(ev$vectors) * sqrt(ev$values))) %*% rbind(cos(t_), sin(t_)) 
  as.data.frame(t(a))
}
coda_ellipse = function(mu, sigma, p, B = ilr_basis(length(mu) + 1)){
  composition(ellipse(mu, sigma, p), B)
}

ggtern() + 
  geom_point(data= as.data.frame(rlrnormal(100, mu, sigma)), aes(x = V1, y = V2, z = V3), alpha = 0.5) +
  geom_path(data = coda_ellipse(mu, sigma, 0.95), aes(x = x1, y = x2, z = x3)) 
```

```{r}
p = ggtern() +
  geom_point(aes(x = 0.4359516, y = 0.4359516, z = 0.1280968), col = 'blue') +
  geom_path(data = coda_ellipse(mu, sigma, 0.95), aes(x = x1, y = x2, z = x3), col = 'blue')
for(i in 45){
  p = p + 
    geom_point(data = M1[i,], aes(x = x1, y = x2, z = x3), alpha = 0.5) +
    geom_path(data = coda_ellipse(unlist(H1[i,]), M[[2]][,,i], 0.95), aes(x = x1, y = x2, z = x3), alpha = 0.5)
}
p
```

```{r}
p = ggplot() +
  geom_point(aes(x = mu[1], y = mu[2]), col = 'blue') +
  geom_path(data = ellipse(mu, sigma, 0.95), aes(x = V1, y = V2), col = 'blue')
for(i in 45){
  p = p + 
    #geom_point(data = coordinates(SL3[i,]), aes(x = V1, y = V2), alpha = 0.5) +
    geom_point(data = H1[i,], aes(x = V1, y = V2), alpha = 0.5) +
    geom_path(data = ellipse(unlist(H1[i,]), M[[2]][,,i], 0.95), aes(x = V1, y = V2), alpha = 0.5)
}
p
```

