---
title: "The logratio-normal-multinomial distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lrnm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = F
)
```

```{r setup, message=FALSE}
library(coda.count)
library(coda.base)
library(ggtern)
theme_set(theme_minimal())
```

The logratio-normal-multinomial (LRNM) distribution is the distribution obtained from the compounding of a normal distribution defined on the Simplex using log-ratio coordinates with the multinomial distribution.

We can build a sample of a LRNM as follows:

* Firstly, for given parameters $\mu$ and $Sigma$ and a certain clr-basis, we generate a log-ratio normally distributed sample :

```{r}
set.seed(1)
N = 200
mu = c(0,1)
sigma = matrix(c(1, -0.8, 
                 -0.8, 1), nrow = 2)
B = ilr_basis(3)
P = as.data.frame(rlrnormal(n = N, mu, sigma, B))
H = coordinates(P, B, label = 'ilr')
```

We can visualise either the compositional sample in a Ternary diagram or their coordinates (with respect basis $B$). 

```{r, warning=FALSE, out.width="45%", fig.width=4, fig.height=3.6, fig.show='hold'}
ggtern() +
  geom_point(data = P, aes(x = V1, y = V2, z = V3))
ggplot() +
  geom_point(data = H, aes(x = ilr1, y = ilr2))
```

* After the logratio-normal random variables are generated, they are used to generate a new sample of counts using the multinomial distribution.

```{r}
SIZE = 8
Y = as.data.frame(rmultinomial(size = SIZE, p = as.matrix(P)))
```

Visualizing a counting sample

```{r, include=FALSE}
Yj = as.data.frame(lapply(Y, jitter, factor = 0.5))
ggtern() +
  geom_point(data = Yj, aes(x = V1, y = V2, z = V3), size = 0.5) +
  scale_T_continuous(minor_breaks = NULL, breaks = 0:SIZE/SIZE, labels = 0:SIZE) +
  scale_L_continuous(minor_breaks = NULL, breaks = 0:SIZE/SIZE, labels = 0:SIZE) +
  scale_R_continuous(minor_breaks = NULL, breaks = 0:SIZE/SIZE, labels = 0:SIZE) 
```

```{r, fig.width=6, fig.height=4.8, fig.align='center'}
key = function(x) paste(x, collapse = '-')
tab = table(apply(Y, 1, key))
Yc = Y[duplicated(Y),]
Yc$freq = as.numeric(tab[match(apply(Yc, 1, key), names(tab))])
lbls = paste(' ', 0:SIZE, ' ')
brks = 0:SIZE/SIZE
ggtern() +
  geom_point(data = Yc, aes(x = V1, y = V2, z = V3, col = freq), size = 3) +
  theme(tern.axis.arrow.show = T,
        tern.axis.title.show = F,
        tern.axis.arrow.text = element_text(face = 'bold')) +
  scale_color_gradient(low = 'blue', high = 'red') +
  scale_T_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  scale_L_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  scale_R_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  labs(col = 'Frequency')
```

For given parameters $mu$ and $\Sigma$ we can calculate the probability mass function

```{r, fig.width=6, fig.height=4.8, fig.align='center'}
X = as.data.frame(simplex_lattice(SIZE, 3))
X$p = apply(X, 1, dlrnm, mu, sigma)
ggtern() +
  geom_point(data = X, aes(x = V1, y = V2, z = V3, col = p), size = 3) +
  theme(tern.axis.arrow.show = T,
        tern.axis.title.show = F,
        tern.axis.arrow.text = element_text(face = 'bold')) +
  scale_color_gradient(low = 'blue', high = 'red') +
  scale_T_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  scale_L_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  scale_R_continuous(minor_breaks = NULL, breaks = brks, labels = lbls) +
  labs(col = 'Probability')
```

```{r}
fitting = fit_lrnm(Y, order = 5)
```

